\documentclass[11pt]{article}

\usepackage{iftex}
\ifPDFTeX
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\fi
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage[numbers]{natbib}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, backgrounds, calc, fit, matrix, positioning}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  citecolor=blue,
  urlcolor=blue
}

\title{KL-TE: Statistically Filtered Hierarchical Clustering with Projected Wald Tests}
\author{Draft workspace}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
A hierarchical tree proposes many candidate splits, but geometric tree cuts do
not say which of those splits are statistically credible. KL-TE is a
statistically filtered hierarchical clustering method for binary,
categorical, and discretized continuous sample-feature data. The method
builds an average-linkage Hamming tree on a working discrete representation,
assigns each internal node feature-wise distributions, and then answers the
node-level split question with two statistical tests. The child-parent edge
test asks whether at least one child truly differs from its parent; it uses a
variance formula that respects the nested parent-child relationship, a
parent-specific summary step, and false-discovery correction across
edges. The sibling test asks whether the two children are genuinely distinct
from one another; it reuses information from the edge stage, compares the
siblings along the main directions of variation seen at the parent, and
applies one global correction for the bias created by testing on a tree built
from the same data before false-discovery correction across focal sibling
pairs. Categorical inputs
are handled through indicator expansion, while continuous inputs enter through
explicit discretization before the same tree-and-test procedure is applied. The
corrected test annotations are converted into clusters by a top-down
pass-through traversal. The manuscript records the implemented method in its
current form and provides the mathematical description needed for empirical
evaluation.
\end{abstract}

\input{sections/introduction/section}
\input{sections/method/section}
\input{sections/implementation/section}
\input{sections/experiments/section}
\input{sections/discussion/section}

\appendix
\input{sections/appendix/projected_wald}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
