% !TEX root = ../../main.tex

\section{Introduction}

Hierarchical clustering, meaning a family of methods that builds nested groups
by repeatedly merging similar samples or sample groups, is attractive when the
analyst wants more than a flat partition. In this setting, a tree is a rooted
diagram of nested groups of
samples. The leaves of the tree are the observed samples themselves. As the
clustering algorithm merges similar samples or groups, it creates internal
nodes, each of which represents the union of all leaves below it. The root is
the final node containing the full data set. Such a representation exposes
coarse-to-fine structure, keeps near-miss cluster boundaries visible, and makes
it possible to discuss not only the final grouping but also the intermediate
grouping decisions that led to it. In practice, however, most hierarchical
workflows still cut the tree using simple geometric rules such as a fixed
height, an inconsistency threshold, or a manually chosen number of clusters.
Those choices are often difficult to defend when the goal is not only
descriptive grouping but also reproducible statistical claims.

KL-TE takes a different view of this tree. An internal node is not merely a
merge produced by the algorithm; it is a candidate decision point. If a parent
node $u$ has children $l(u)$ and $r(u)$, then that parent proposes a possible
split of the data into two child subtrees. The central question is whether this
proposed split should survive in the final partition. A linkage tree alone
cannot answer that question. Here, a linkage tree means the hierarchy returned
by the agglomerative merging rule itself. It proposes candidate splits, but it
does not tell
us whether at least one child genuinely departs from the parent's feature
distribution, whether the two children are actually distinct enough to
represent separate clusters, or whether scanning the entire tree has produced
false positives.

The method therefore requires two kinds of evidence. First, a split should not
survive unless at least one child meaningfully departs from its ancestral
context. Second, even that is not enough: a parent can contain one drifting
branch or a mild asymmetry without supporting two stable child clusters, so the
children themselves must also be distinguishable from one another. This is why
the method uses both a child-parent edge test and a sibling test. The final
partition keeps a split only when both questions are answered favorably after
multiple-testing correction, that is, an adjustment for the fact that many
candidate splits are tested across the same hierarchy.

Four inferential complications shape the design. The child is a subset of its
parent, so the edge comparison is nested rather than an ordinary two-sample
test. The edge evidence is high-dimensional and correlated, so a simple sum of
squared feature scores would overcount repeated or noisy directions. Many candidate edges
are examined across the tree, so valid local $p$-values still require control
for many tests. Finally, sibling comparisons are evaluated on a hierarchy
constructed from the same data, so the raw sibling statistics are selected and
inflated. KL-TE is one response to these four problems.

KL-TE works on a discrete feature representation. Binary data
enter directly. Categorical data are represented through one-hot or indicator-expanded
binary features. Continuous data are discretized before the same tree-and-test
procedure is applied. Once the representation is fixed, the inferential objects
are no longer individual samples but subtrees, so each node carries a
feature-wise distribution estimated from its descendant leaves. Statistical
test annotations are computed before the actual tree walk that creates cluster
assignments. This separation between evidence calculation and final traversal
is useful both computationally and scientifically: the statistical support for
each potential split is stored explicitly and can be audited independently of
the resulting partition.

The presentation follows that decision problem in order. It first constructs the
tree and node distributions that define the candidate splits. It then asks
whether a child departs from its parent, explains why that test needs a nested
variance formula because the child group is contained inside the parent group,
and uses a parent-local summary of the feature space, meaning a lower-dimensional
description built from the variation visible below that parent. It then applies
hierarchical multiple-testing control across the tree. Next it asks whether the
two children under the same parent are genuinely distinct, and explains why
that sibling comparison needs a post-selection calibration step, meaning a
correction for the bias introduced when the same data are used both to choose a
split and to test it. Only after both evidence tables are in place does the
final tree traversal, that is, the top-down walk that decides where splitting
stops, convert them into clusters.
