% !TEX root = ../main.tex

\section{Experimental Plan}

This section is a forward-looking placeholder. Before the method can support
publication-quality inferential claims, the manuscript should cover at least
the following empirical blocks.

\subsection{Null calibration}

The first requirement is a null calibration study on binary data with no true
cluster structure. The goal is to measure how often the child-parent edge test
and the sibling test reject under scenarios where they should not. This should
be reported both at the
individual-test level and at the resulting number-of-clusters level. The latter
is critical because a method can have mild per-test inflation yet still produce
severe over-splitting once the tree walk is applied.

\subsection{Power under planted structure}

The second requirement is a controlled power study with planted binary subtrees.
This should vary:

\begin{itemize}[leftmargin=1.5em]
  \item sample size per cluster,
  \item effect size in feature rates,
  \item number of informative features,
  \item tree imbalance, and
  \item background noise level.
\end{itemize}

Power should be measured both for edge recovery and for final cluster recovery.
Adjusted Rand Index and exact split recovery are both useful endpoints.

\subsection{Robustness analyses}

Because the manuscript now presents a single implemented method, the relevant
empirical question is robustness of that pipeline. The most useful stress tests
are:

\begin{itemize}[leftmargin=1.5em]
  \item sensitivity to small child leaf counts in the child-parent edge test,
  \item sensitivity to sparse high-dimensional feature matrices,
  \item sensitivity to tree imbalance and shallow-versus-deep signal placement, and
  \item sensitivity of the final partition to the pass-through rule.
\end{itemize}

This section should eventually explain where the implemented method is well
calibrated, where it becomes under-powered, and where it tends to over-split or
under-split.

\subsection{Real-data benchmarks}

The repository already contains benchmark infrastructure and reporting code. A
manuscript-quality benchmark section should align those experiments with the
method claims in the text, focusing on:

\begin{itemize}[leftmargin=1.5em]
  \item synthetic binary benchmarks with known labels,
  \item sensitivity to over-splitting and under-splitting,
  \item comparison to standard clustering baselines, and
  \item qualitative inspection of annotated trees.
\end{itemize}
