% !TEX root = ../main.tex

\section{Method}

\subsection{Data representation and notation}

Let $X$ denote the working sample representation after any preprocessing. In
the direct binary or indicator case,
$X \in \{0,1\}^{n \times p}$ with rows $X_i \in \{0,1\}^p$. Categorical and
continuous inputs are handled through the representation maps described later
in this section. The method builds a rooted agglomerative tree
$\mathcal{T} = (V,E)$ using Hamming distance between samples and average
linkage. Leaves are in one-to-one correspondence with the original samples.

For each node $u \in V$, let $L(u)$ denote its descendant leaf set and let
$n_u = |L(u)|$. For binary or indicator coordinates, the node distribution is
the feature-wise mean vector
\begin{equation}
\theta_u = \frac{1}{n_u}\sum_{i \in L(u)} X_i \in [0,1]^p.
\end{equation}
If $u$ has children $l$ and $r$, then
\begin{equation}
n_u = n_l + n_r,
\qquad
\theta_u = \frac{n_l \theta_l + n_r \theta_r}{n_u}.
\end{equation}
The decomposition goal is to identify a set of terminal tree nodes whose leaf
sets partition the samples and whose defining splits are supported by the two
statistical gates below. These formulas cover binary data directly and also the
indicator representations used for categorical and discretized continuous data.

\subsection{Gate 2: child-parent divergence test}

For each edge $(u,c) \in E$, Gate 2 tests whether child $c$ carries signal
beyond its parent $u$. The coordinate-wise null is
$H_{0,j}^{\mathrm{edge}}(u,c): \theta_{c,j} = \theta_{u,j}$. Because $c$ is
nested inside $u$, the child and parent estimators are not independent. In the
binary or indicator representation,
\begin{equation}
\mathrm{Var}(\hat{\theta}_{c,j} - \hat{\theta}_{u,j})
=
\theta_{u,j}(1-\theta_{u,j})
\left(\frac{1}{n_c} - \frac{1}{n_u}\right),
\end{equation}
which is the variance expression used in the code. The standardized edge vector
is therefore
\begin{equation}
z_{u \to c,j}^{\mathrm{edge}}
=
\frac{\theta_{c,j} - \theta_{u,j}}
{\sqrt{\theta_{u,j}(1-\theta_{u,j})\left(\frac{1}{n_c} - \frac{1}{n_u}\right)}}.
\end{equation}
In the current parameterization there is no Felsenstein branch-length
rescaling, so the variance above is used as written.

Gate 2 uses a parent-local spectral projection. For each parent node $u$, the
implementation forms a local matrix $M_u$ whose rows are:
\begin{enumerate}[leftmargin=1.5em]
  \item all descendant leaf vectors $X_i$ for $i \in L(u)$, and
  \item all descendant internal-node distributions $\theta_v$ for proper
  descendants $v$ of $u$.
\end{enumerate}
This second block is included because internal distributions are part of the
current spectral construction. After removing constant features, let $d_u$ be
the number of active coordinates and let $m_u$ be the number of rows in $M_u$.
The code computes the correlation matrix
$C_u \in \mathbb{R}^{d_u \times d_u}$ and its eigen-decomposition
\begin{equation}
C_u = V_u \Lambda_u V_u^\top,
\qquad
\Lambda_u = \mathrm{diag}(\lambda_{u,1},\dots,\lambda_{u,d_u}),
\qquad
\lambda_{u,1} \ge \cdots \ge \lambda_{u,d_u} \ge 0.
\end{equation}

The spectral dimension is the Marchenko--Pastur signal count
\citep{marchenko1967distribution}. With
\begin{equation}
\hat{\sigma}_u^2 = \mathrm{median}\{\lambda_{u,i} : \lambda_{u,i} > 0\},
\qquad
b_u^{\mathrm{MP}} = \hat{\sigma}_u^2 \left(1 + \sqrt{\frac{d_u}{m_u}}\right)^2,
\end{equation}
the projection dimension is
\begin{equation}
k_u^{\mathrm{edge}}
=
\max\left\{
2,\;
\#\{i : \lambda_{u,i} > b_u^{\mathrm{MP}}\}
\right\},
\end{equation}
followed by the implementation cap $k_u^{\mathrm{edge}} \le d_u$. The floor of
$2$ is part of the current parameterization.

Let $v_{u,1},\dots,v_{u,k_u^{\mathrm{edge}}}$ denote the leading eigenvectors.
Because eigenvalue whitening is enabled, the Gate 2 test statistic is
\begin{equation}
T_{u \to c}^{\mathrm{edge}}
=
\sum_{i=1}^{k_u^{\mathrm{edge}}}
\frac{(v_{u,i}^\top z_{u \to c}^{\mathrm{edge}})^2}{\lambda_{u,i}}.
\end{equation}
Under the local correlation-model approximation used by the implementation,
\begin{equation}
T_{u \to c}^{\mathrm{edge}} \sim \chi^2_{k_u^{\mathrm{edge}}},
\qquad
p_{u \to c}^{\mathrm{edge}}
=
\Pr\!\left(
\chi^2_{k_u^{\mathrm{edge}}}
\ge
T_{u \to c}^{\mathrm{edge}}
\right).
\end{equation}
These $p$-values are computed for every edge in the tree.

\subsection{Tree-aware multiple testing for Gate 2}

Gate 2 is not used edge-by-edge in isolation. The pipeline applies the
TreeBH procedure \citep{bogomolov2021hypotheses} at level
$\alpha_{\mathrm{edge}} = 0.01$. Edge hypotheses are grouped into families by
parent node. If $F(u)$ denotes the family of edges from parent $u$ to its
children, then the family-specific level used by the implementation is
\begin{equation}
\alpha_u
=
\alpha_{\mathrm{edge}}
\prod_{a \in \mathrm{Anc}(u)}
\frac{R_a}{|F(a)|},
\end{equation}
where $R_a$ is the number of rejections already made in the ancestor family
$F(a)$. Root families start at $\alpha_{\mathrm{edge}}$. Within each family,
ordinary Benjamini--Hochberg is applied at level $\alpha_u$. Let
$S^{\mathrm{edge}}(c) \in \{0,1\}$ denote the resulting significance indicator
for child $c$.

\subsection{Gate 3: cousin-adjusted sibling test}

For each parent $u$ with exactly two children $l(u)$ and $r(u)$, Gate 3 tests
the sibling null
\begin{equation}
H_0^{\mathrm{sib}}(u): \theta_{l(u)} = \theta_{r(u)}.
\end{equation}
For binary or indicator coordinates, the pooled null proportion is
\begin{equation}
\bar{\theta}_{u,j}
=
\frac{n_{l(u)}\theta_{l(u),j} + n_{r(u)}\theta_{r(u),j}}
{n_{l(u)} + n_{r(u)}}.
\end{equation}
The standardized sibling vector is
\begin{equation}
z_{u,j}^{\mathrm{sib}}
=
\frac{\theta_{l(u),j} - \theta_{r(u),j}}
{\sqrt{
\bar{\theta}_{u,j}(1-\bar{\theta}_{u,j})
\left(\frac{1}{n_{l(u)}} + \frac{1}{n_{r(u)}}\right)
}}.
\end{equation}
Again, the current parameterization does not apply Felsenstein variance
inflation.

Unlike Gate 2, Gate 3 does not use spectral decomposition. Instead it uses an
orthonormal random projection with a Johnson--Lindenstrauss-style dimension
rule \citep{johnson1984extensions}. Let $k_{\min}$ be the run-level minimum
projection dimension computed once from the global effective rank of the full
dataset,
\begin{equation}
k_{\min}
=
\max\left\{
2,\;
\min\left(\left\lceil \mathrm{erank}(C_{\mathrm{full}}) \right\rceil, 20\right)
\right\},
\end{equation}
where $\mathrm{erank}(C_{\mathrm{full}})=
\exp\!\left(-\sum_i p_i \log p_i\right)$ for
$p_i = \lambda_i / \sum_j \lambda_j$. For sibling pair $u$, the projection
dimension is
\begin{equation}
k_u^{\mathrm{sib}}
=
\min\left\{
p,\;
\max\left(\mathrm{JL}(n_{l(u)} + n_{r(u)}, p; \varepsilon=0.3),\; k_{\min}\right)
\right\},
\end{equation}
with the additional implementation cap
$k_u^{\mathrm{sib}} \le n_{l(u)} + n_{r(u)}$ when $p \ge 4(n_{l(u)} + n_{r(u)})$.
If $R_u \in \mathbb{R}^{k_u^{\mathrm{sib}} \times p}$ has orthonormal rows,
the raw sibling statistic is
\begin{equation}
T_u^{\mathrm{raw}}
=
\|R_u z_u^{\mathrm{sib}}\|_2^2,
\qquad
p_u^{\mathrm{raw}}
=
\Pr\!\left(\chi^2_{k_u^{\mathrm{sib}}} \ge T_u^{\mathrm{raw}}\right).
\end{equation}

The sibling calibration is the cousin-adjusted Wald model. The code computes
raw sibling statistics for every parent with exactly two children, then splits
them into two groups:
\begin{enumerate}[leftmargin=1.5em]
  \item \emph{null-like pairs}, where neither child is Gate 2 significant, and
  \item \emph{focal pairs}, where at least one child is Gate 2 significant.
\end{enumerate}
A null-like pair is reclassified as focal if its raw sibling
$p$-value satisfies $p_u^{\mathrm{raw}} \le 10^{-3}$. This is an explicit
self-consistency rule in the current implementation.

For each calibration pair $q$, define the inflation ratio
\begin{equation}
r_q = \frac{T_q^{\mathrm{raw}}}{k_q^{\mathrm{sib}}}.
\end{equation}
Let $b_q$ denote the sum of the two child branch lengths under parent $q$, and
let $n_q$ denote the parent leaf count. The post-selection inflation model is
\begin{equation}
\log r_q = \beta_0 + \beta_1 \log b_q + \beta_2 \log n_q + \varepsilon_q.
\end{equation}
The fitting rule is exactly tiered:
\begin{enumerate}[leftmargin=1.5em]
  \item if fewer than $3$ valid null-like pairs are available, no calibration
  is applied and $\hat{c}_u = 1$;
  \item if $3$ or $4$ valid null-like pairs are available, the method uses the
  global median $\hat{c}_u = \mathrm{median}(r_q)$;
  \item if at least $5$ valid null-like pairs are available, the regression is
  fit, but the method still falls back to the global median if the fitted
  $R^2 < 0.05$.
\end{enumerate}
For a focal pair $u$, the predicted inflation factor is
\begin{equation}
\hat{c}_u
=
\exp(\beta_0 + \beta_1 \log b_u + \beta_2 \log n_u),
\end{equation}
then clamped to the interval
$[1,\max_q r_q]$. If $b_u \le 0$, the code falls back to the global median
inflation factor rather than extrapolating. The adjusted sibling test is
\begin{equation}
T_u^{\mathrm{adj}} = \frac{T_u^{\mathrm{raw}}}{\hat{c}_u},
\qquad
p_u^{\mathrm{sib}}
=
\Pr\!\left(\chi^2_{k_u^{\mathrm{sib}}} \ge T_u^{\mathrm{adj}}\right).
\end{equation}

The sibling multiplicity correction is ordinary
Benjamini--Hochberg \citep{benjamini1995controlling} at
$\alpha_{\mathrm{sib}} = 0.01$, applied only across focal sibling pairs. Let
$S^{\mathrm{sib}}(u) \in \{0,1\}$ denote the resulting indicator that parent
$u$ supports a significant sibling split.

\subsection{Categorical and continuous representations}

Our method is not limited to direct Bernoulli features. A categorical feature
with values $C_{ij} \in \{0,\dots,K-1\}$ can be represented through one-hot
indicators
\begin{equation}
\tilde{X}_{i,j,k} = \mathbf{1}\{C_{ij} = k\},
\qquad
k = 0,\dots,K-1,
\end{equation}
so that the Bernoulli/indicator formulas above apply directly on the expanded
feature space. This is the practical route used by much of the benchmark and
data-generation code.

The gate code also supports native categorical node distributions
$\Theta_u \in [0,1]^{d \times K}$ with row sums equal to one. In Gate 2, the
current implementation forms elementwise standardized scores
\begin{equation}
z_{u \to c,jk}^{\mathrm{cat}}
=
\frac{\Theta_{c,jk} - \Theta_{u,jk}}
{\sqrt{
\Theta_{u,jk}(1-\Theta_{u,jk})
\left(\frac{1}{n_c} - \frac{1}{n_u}\right)
}},
\end{equation}
then drops the last category of each feature before projection so that only
$K-1$ coordinates per feature remain. This is a simplex-aware reduction, but it
still uses a diagonal variance approximation in Gate 2.

In Gate 3, native categorical inputs are handled with a stronger multinomial
covariance model. Let
$\bar{\Theta}_{u,j}^{-} \in \mathbb{R}^{K-1}$ be the pooled sibling proportion
for feature $j$ after dropping the last category, and let
$\Delta_{u,j}^{-} \in \mathbb{R}^{K-1}$ be the corresponding left-right
difference. With effective sample size
\begin{equation}
n_{\mathrm{eff}} = \frac{n_{l(u)} n_{r(u)}}{n_{l(u)} + n_{r(u)}},
\end{equation}
the per-feature covariance block is
\begin{equation}
\Sigma_{u,j}
=
\frac{
\mathrm{diag}(\bar{\Theta}_{u,j}^{-})
-
\bar{\Theta}_{u,j}^{-}(\bar{\Theta}_{u,j}^{-})^\top
}{n_{\mathrm{eff}}}
\;+\;
\lambda I,
\end{equation}
where $\lambda$ is a small ridge term for numerical stability. If
$L_{u,j}L_{u,j}^\top = \Sigma_{u,j}$ is the Cholesky factorization, the
whitened categorical sibling vector is built featurewise as
\begin{equation}
z_{u,j}^{\mathrm{cat}} = L_{u,j}^{-1}\Delta_{u,j}^{-},
\end{equation}
then concatenated across features and passed through the same projected
Wald, cousin-adjusted, and BH-corrected sibling pipeline described above.

Continuous inputs enter through an explicit preprocessing map into a discrete
representation. The two current routes are median binarization,
\begin{equation}
X_{ij}^{\mathrm{bin}}
=
\mathbf{1}\!\left\{
W_{ij} > \mathrm{median}(W_{\cdot j})
\right\},
\end{equation}
and quantile discretization followed by one-hot expansion,
\begin{equation}
C_{ij}
=
\sum_{q=1}^{K-1} \mathbf{1}\{W_{ij} > t_{jq}\},
\qquad
t_{jq} = Q_{q/K}(W_{\cdot j}),
\end{equation}
with the same indicator map
$\tilde{X}_{i,j,k} = \mathbf{1}\{C_{ij} = k\}$. After this representation step,
the same hierarchy construction and gate logic are used.

\subsection{Recursive decomposition rule}

The final clusters are produced by a top-down traversal that reads the two gate
annotation tables. For an internal node $u$ with children $l(u)$ and $r(u)$,
the split rule is
\begin{equation}
\mathrm{Split}(u) = 1
\end{equation}
if and only if all three conditions hold:
\begin{enumerate}[leftmargin=1.5em]
  \item $u$ has exactly two children;
  \item $S^{\mathrm{edge}}(l(u)) = 1$ or $S^{\mathrm{edge}}(r(u)) = 1$;
  \item $S^{\mathrm{sib}}(u) = 1$.
\end{enumerate}
If these conditions hold, traversal continues into both children. Otherwise the
method would ordinarily stop and declare $u$ a terminal cluster.

However, the implementation also enables passthrough traversal. Let
$D^{\mathrm{sib}}(u)$ be the indicator that some proper descendant of $u$ has a
significant non-skipped sibling split. Then the traversal still descends into
both children when
\begin{equation}
\mathrm{Split}(u) = 0,
\qquad
u \text{ has exactly two children},
\qquad
S^{\mathrm{edge}}(l(u)) \vee S^{\mathrm{edge}}(r(u)) = 1,
\qquad
D^{\mathrm{sib}}(u) = 1.
\end{equation}
This is the passthrough rule used to prevent a shallow Gate 3 failure from
masking deeper structure. Because post-hoc merge is off, the terminal nodes
returned by this traversal define the final cluster partition directly. Because
edge calibration is off, Gate 2 uses only the TreeBH-corrected projected Wald
evidence described above.
