% !TEX root = ../../main.tex

\subsection{Data Representation and Subtree Objects}

\Cref{tab:notation} fixes the global symbols used in the rest of the
method. This subsection introduces the tree objects in plain language before
writing them formally, because the statistical tests operate on subtree
summaries rather than directly on individual samples. All supported inputs are
first mapped into one working discrete sample-feature matrix $X$, and the
method then builds a rooted agglomerative tree $\mathcal{T} = (V,E)$ using
Hamming distance and average linkage.

For a non-expert reader, the important point is that the tree is a
bookkeeping device for nested groups of samples. Each leaf corresponds to one
observed sample. Each internal node is created by one agglomerative merge and
therefore stands for a group of leaves rather than a new observation. The root
is the unique node that contains every sample. In that sense, internal nodes
represent candidate subpopulations, while leaves represent the original data
points.

The edges of the tree record immediate containment relations between these
groups. If $(u,c) \in E$, then $c$ is a child subtree sitting directly inside
the larger parent subtree $u$. Equivalently, every leaf below $c$ is also
below $u$, and there is no intermediate node between them. This interpretation
is important because the edge test is attached to exactly that parent-child
relation: it asks whether the child subtree distribution departs from the
background defined by its parent. The sibling test then uses the two child
edges under the same parent to ask whether those child subtrees should remain
separate.

In the direct binary or indicator case, $X \in \{0,1\}^{n \times p}$ with rows
$X_i \in \{0,1\}^p$. The leaf set of $\mathcal{T}$ consists of the original
samples, and each agglomerative merge creates one new internal node whose
descendant set is the union of the two merged child subtrees. Starting from $n$
singleton leaves and merging until all samples belong to one connected
hierarchy yields a rooted binary tree with
\begin{equation}
|V| = 2n-1,
\qquad
|E| = 2n-2,
\end{equation}
consisting of $n$ leaves, $n-1$ internal nodes, and one root. The root
represents the full sample set, and each internal node defines one candidate
split location that may or may not survive the later statistical filtering.
Formally, if $u \in V$ is any node, then its associated subtree object is the
set of descendant leaves
\begin{equation}
L(u) = \{ i \in \{1,\dots,n\} : \text{leaf } L_i \text{ lies below } u \}.
\end{equation}
The subtree size is
\begin{equation}
n_u = |L(u)|.
\end{equation}
If $c$ is a child of $u$, then the edge $(u,c)$ represents the immediate set
inclusion
\begin{equation}
L(c) \subset L(u),
\end{equation}
and if $u$ has children $l(u)$ and $r(u)$, then the binary-tree structure gives
the disjoint union identity
\begin{equation}
L(u) = L\bigl(l(u)\bigr) \cup L\bigl(r(u)\bigr),
\qquad
L\bigl(l(u)\bigr) \cap L\bigl(r(u)\bigr) = \varnothing.
\end{equation}

Once that tree is fixed, the inferential problem is about subtrees rather than
individual samples. The method therefore replaces each node by an empirical
feature distribution over its descendant leaves. This is the object that enters
both tests: the edge test compares a child distribution to its parent's
distribution, and the sibling test compares the two child distributions under
the same parent. The raw leaf matrix still matters through the descendant sets
and local spectral matrices, but the primary inferential units are node-level
distributions.

For binary or indicator coordinates, the node distribution is the feature-wise
mean over the descendant leaves,
\begin{equation}
\theta_u = \frac{1}{n_u}\sum_{i \in L(u)} X_i \in [0,1]^p.
\end{equation}
Thus $\theta_u$ is not a parameter attached to a single sample. It is the
empirical summary of the whole subtree rooted at $u$: each coordinate
$\theta_{u,j}$ is the fraction of descendant leaves in that subtree whose
$j$th working feature equals $1$.
If $u$ has children $l$ and $r$, then
\begin{equation}
n_u = n_{l(u)} + n_{r(u)},
\qquad
\theta_u = \frac{n_{l(u)} \theta_{l(u)} + n_{r(u)} \theta_{r(u)}}{n_u}.
\end{equation}
The decomposition goal is to identify a set of terminal tree nodes whose leaf
sets partition the samples and whose defining splits are supported by the two
statistical tests below. These formulas cover binary data directly and also the
indicator representations used for categorical and discretized continuous data.

Categorical features with values $C_{ij} \in \{0,\dots,K-1\}$ are represented
through one-hot indicators
\begin{equation}
\tilde{X}_{i,j,k} = \mathbf{1}\{C_{ij} = k\},
\qquad
k = 0,\dots,K-1,
\end{equation}
so that the same Bernoulli or indicator formulas apply on the expanded feature
space. Continuous inputs enter through an explicit preprocessing map into a
discrete representation. One simple option is median binarization,
\begin{equation}
X_{ij}^{\mathrm{bin}}
=
\mathbf{1}\!\left\{
W_{ij} > \mathrm{median}(W_{\cdot j})
\right\},
\end{equation}
while a more granular option is quantile discretization followed by one-hot
expansion,
\begin{equation}
C_{ij}
=
\sum_{q=1}^{K-1} \mathbf{1}\{W_{ij} > t_{jq}\},
\qquad
t_{jq} = Q_{q/K}(W_{\cdot j}),
\end{equation}
with the same indicator map
$\tilde{X}_{i,j,k} = \mathbf{1}\{C_{ij} = k\}$. After this representation step,
the same tree construction and testing logic are used for all supported input
types.
