% !TEX root = ../../main.tex

\subsection{Sibling Test: Are the Two Children Distinct?}

The sibling test is the second inferential step. For each parent $u$ with
exactly two children $l(u)$ and $r(u)$, it asks whether the two children are
genuinely different from one another rather than merely both differing from the
parent. The edge test alone is not enough to justify keeping a split. A parent
can contain one deviating branch, or a weak left-right asymmetry, without
supporting two stable child clusters. The sibling test supplies that missing
question: once a branch has left its ancestral context, are the two children
actually distinct enough to become separate clusters?
Because the hierarchy is a rooted binary tree, this sibling relation is always
simple: every non-root node has exactly one parent and therefore at most one
sibling, namely the other child of that same parent. The sibling test is thus
always a single pairwise comparison, not a comparison against a larger sibling
set.

\paragraph{Sibling null and pooled baseline.}
Formally, it tests the sibling null
\begin{equation}
H_0^{\mathrm{sib}}(u): \theta_{l(u)} = \theta_{r(u)}.
\end{equation}
For binary or indicator coordinates, the pooled null proportion is
\begin{equation}
\bar{\theta}_{u,j}
=
\frac{n_{l(u)}\theta_{l(u),j} + n_{r(u)}\theta_{r(u),j}}
{n_{l(u)} + n_{r(u)}}.
\end{equation}

\paragraph{Standardized sibling contrast.}
Because the two children are disjoint subtrees under the same parent, the
coordinatewise variance takes the usual two-sample Bernoulli form rather than
the nested child-parent form used in the edge test. The standardized sibling
vector is
\begin{equation}
z_{u,j}^{\mathrm{sib}}
=
\frac{\theta_{l(u),j} - \theta_{r(u),j}}
{\sqrt{
\bar{\theta}_{u,j}(1-\bar{\theta}_{u,j})
\left(\frac{1}{n_{l(u)}} + \frac{1}{n_{r(u)}}\right)
}}.
\end{equation}
The sibling test then asks how to combine this multi-feature contrast into one
summary score. The method reuses the spectral information already
computed during the child-parent stage rather than drawing a fresh random
projection from scratch. For a binary parent $u$ with children $l(u)$ and
$r(u)$, let $k_{l(u)}^{\mathrm{edge}}$ and $k_{r(u)}^{\mathrm{edge}}$ denote
the positive edge-stage spectral dimensions of the two children. The sibling
dimension used in code is
\begin{equation}
k_u^{\mathrm{sib}}
=
\operatorname{round}
\!\left(
\sqrt{
k_{l(u)}^{\mathrm{edge}}
k_{r(u)}^{\mathrm{edge}}
}
\right).
\end{equation}
If exactly one child has a positive edge-stage spectral dimension, that value
is reused directly; if neither child has a positive dimension, the sibling
test is skipped and the node is treated conservatively as a merge. This
standardized vector answers the first sibling question: how large is the
left-right contrast after accounting for feature-specific sampling noise? The
next question is how to combine those coordinatewise contrasts into one summary
score.

\paragraph{Projected sibling statistic.}
The sibling test then combines this multi-feature contrast into a single
summary score while preserving the low-dimensional structure already learned at
the parent during the edge stage.

The projection basis starts with the parent node, not a fresh random draw. The
first available parent PCA directions, where PCA refers to principal-component
directions summarizing the main variation at that parent, are used up to
$k_u^{\mathrm{sib}}$. If the requested sibling dimension exceeds the available
parent PCA rank, the remaining rows are filled with orthonormal padding
directions. Version-specific basis-construction details are summarized in
\Cref{sec:software-implementation}.

If $P_u \in \mathbb{R}^{k_u^{\mathrm{sib}} \times p}$ denotes this sibling
projection basis, the raw sibling statistic is
\begin{equation}
T_u^{\mathrm{raw}}
=
\|P_u z_u^{\mathrm{sib}}\|_2^2.
\end{equation}

\paragraph{Reference-law calibration.}
The default p-value calculation is deliberately conservative. After
projection, the method asks whether the observed squared contrast is larger
than would be expected from ordinary within-parent sampling noise, while still
allowing different projected directions to have different variances. In the
method described here, this is done with a Satterthwaite moment match, meaning
an approximation that replaces the exact null law by a scaled chi-square law
with the same first two moments, for the
PCA-derived part of the test and plain $\chi^2(1)$ contributions for any added
orthogonal-complement directions.

\paragraph{Worked sibling example.}
A small numerical example helps separate three ideas that are easy to conflate:
projection, per-component whitening, and Satterthwaite calibration. Suppose
the standardized sibling difference vector is
\begin{equation}
z_u^{\mathrm{sib}} =
\begin{bmatrix}
2.0\\
-1.0\\
0.5\\
1.0
\end{bmatrix},
\qquad
P_u =
\begin{bmatrix}
1&0&0&0\\
0&1&0&0\\
0&0&1&0
\end{bmatrix}.
\end{equation}
Then the projected contrast is
\begin{equation}
P_u z_u^{\mathrm{sib}} =
\begin{bmatrix}
2.0\\
-1.0\\
0.5
\end{bmatrix}.
\end{equation}
Without any PCA variance correction, the raw quadratic form is
\begin{equation}
T_u^{\mathrm{raw}}
=
\|P_u z_u^{\mathrm{sib}}\|_2^2
=
2.0^2 + (-1.0)^2 + 0.5^2
=
5.25.
\end{equation}
If those three projected coordinates were all unit-variance Gaussian noise,
this statistic would be compared directly to $\chi^2_3$. Now suppose the
first two projected directions are parent PCA directions with eigenvalues
$\lambda_1 = 4$ and $\lambda_2 = 1$, while the third coordinate is an added
unit-variance direction. In matrix form,
\begin{equation}
\Lambda =
\begin{bmatrix}
4&0\\
0&1
\end{bmatrix},
\qquad
\Lambda^{-1} =
\begin{bmatrix}
1/4&0\\
0&1
\end{bmatrix}.
\end{equation}
Per-component whitening would then use
\begin{equation}
T_u^{\mathrm{white}}
=
\begin{bmatrix}
2.0 & -1.0
\end{bmatrix}
\Lambda^{-1}
\begin{bmatrix}
2.0\\
-1.0
\end{bmatrix}
+
0.5^2
=
\frac{2.0^2}{4}
+
\frac{(-1.0)^2}{1}
+
0.5^2
=
2.25.
\end{equation}
The first direction contributes less after whitening because a contrast of
size $2.0$ is less surprising along a direction whose null variance is already
$4$. The implemented sibling default instead keeps the raw quadratic form
$T_u^{\mathrm{raw}} = 5.25$ and changes the reference law. The PCA part is a
weighted sum of $\chi^2(1)$ terms with weights $4$ and $1$, and the added
direction contributes one ordinary $\chi^2(1)$ term. Hence
\begin{equation}
\mathbb{E}[T_u^{\mathrm{raw}}] = 4 + 1 + 1 = 6,
\qquad
\mathrm{Var}(T_u^{\mathrm{raw}}) = 2(4^2 + 1^2 + 1^2) = 36.
\end{equation}
Matching those moments to $c\chi^2(\nu)$ gives
\begin{equation}
c = \frac{36}{2 \cdot 6} = 3,
\qquad
\nu = \frac{2 \cdot 6^2}{36} = 2,
\end{equation}
so the calibration becomes
\begin{equation}
T_u^{\mathrm{raw}} \approx 3\chi^2(2),
\qquad
p_u^{\mathrm{sib}}
=
\Pr\!\left(
\chi^2_2 \ge \frac{5.25}{3}
\right).
\end{equation}
This is the practical difference between whitening and Satterthwaite
calibration: whitening rescales the coordinates before summing, whereas
Satterthwaite keeps the raw quadratic form and adjusts the reference
distribution instead. \Cref{app:projected-wald-example} expands this
example in full matrix form and shows the corresponding plain, whitened, and
Satterthwaite-calibrated calculations side by side.

\paragraph{Post-selection inflation.}
Raw sibling $p$-values are still not enough on their own because the sibling
pair being tested is not fixed in advance. The same data are used first to
construct the hierarchy and thereby expose pairs of child subtrees that already
look well separated, and then again to test whether those selected siblings are
significantly different. This data reuse creates \emph{post-selection
inflation}. Even under a null setting with no true cluster structure, the tree
construction step tends to present unusually separated sibling pairs simply
because it was built from the observed dissimilarities. The sibling statistic
is therefore evaluated conditionally on a favorable selection event rather than
under the unconditional null distribution for a pre-specified pair. The
practical consequence is anti-conservative inference: the observed sibling
difference vector $z_u^{\mathrm{sib}}$ is systematically larger than the naive
reference law expects, so raw $p$-values are too small and false split
decisions become too common. Empirical null simulations confirm this effect:
without correction, the Wald sibling test produces Type~I error rates of
approximately 39\% at $\alpha = 0.05$.

\paragraph{Calibration via continuous edge-weight deflation.}
The sibling calibration then corrects for this extra optimism by estimating a
global adjustment factor from the sibling statistics themselves, using the
edge-test $p$-values as continuous weights. For every binary parent $q$, let
$T_q^{\mathrm{raw}}$ be the raw sibling statistic and let $\nu_q$ be the
effective sibling degrees of freedom returned by the projected-Wald kernel. The
inflation ratio is
\begin{equation}
r_q = \frac{T_q^{\mathrm{raw}}}{\nu_q}.
\end{equation}
Under a pure-null calibration model, these ratios should fluctuate around a
common inflation level. The weighting rule is
\begin{equation}
w_q = \min\!\left(p^{\mathrm{edge}}_{l(q)},\;
                    p^{\mathrm{edge}}_{r(q)}\right),
\end{equation}
where $p^{\mathrm{edge}}_{l(q)}$ and $p^{\mathrm{edge}}_{r(q)}$ are the
Benjamini--Hochberg-corrected edge-test $p$-values of the two children. Pairs where neither
child is edge-significant receive large weights and dominate the estimate,
while pairs with genuine cluster signal receive small weights and are
down-weighted. For this calibration step, child edges that were blocked or never tested
are treated as $p=1$ for this calibration weight only. The global inflation
factor is then the
weighted mean
\begin{equation}\label{eq:chat-global}
\hat{c}
=
\frac{\sum_q w_q \, r_q}{\sum_q w_q},
\qquad
\text{clamped to } [1,\;\max_q r_q].
\end{equation}

\paragraph{Why continuous weights, not a binary oracle.}
An earlier design classified sibling pairs into discrete \emph{null-like}
(neither child edge-significant) and \emph{focal} groups, then fitted the
calibration model only on null-like pairs.  This approach has a circular
dependency: on null data, the edge test itself is inflated by post-selection
bias and declares essentially every edge significant---leaving zero null-like
calibration pairs and forcing $\hat{c}=1$ (no correction).  Empirical tracing
on pure-noise data ($n=100$, $p=50$, $K_{\mathrm{true}}=1$) confirmed this
failure: the edge test marked 198 of 199 edges significant, the calibration
received zero null-like pairs, and the method returned $K=17$. Continuous
weighting avoids this dead zone: even when all edge $p$-values are similarly
small, the weights are approximately equal, and the weighted mean of $r_q$
still recovers the true inflation $\hat{c} \approx 3$.

\paragraph{Runtime deflation and sibling false-discovery-rate correction.}
The method first fits one global weighted-mean inflation level
$\hat{c}_{\mathrm{global}}$. It then localizes that correction in the
decomposition-derived sibling structural dimension
$k_q^{\mathrm{struct}}$. This structural dimension is the same quantity used to
choose the sibling projection dimension. As a result, the
calibration neighborhood is defined in the same geometric coordinate system as
the test itself.

For all valid sibling pairs in the calibration pool, define
\[
\ell_q = \log\!\bigl(\max(k_q^{\mathrm{struct}}, 1)\bigr),
\qquad
\bar{\ell}
=
\frac{\sum_q w_q \ell_q}{\sum_q w_q},
\qquad
h
=
\sqrt{\frac{\sum_q w_q (\ell_q - \bar{\ell})^2}{\sum_q w_q}}.
\]
So $\exp(\bar{\ell})$ is the edge-weighted geometric-mean structural
dimension, and $h$ is the edge-weighted standard deviation in log-$k$ space.
For a focal sibling pair $u$, the runtime assigns each calibration pair $q$
the local similarity weight
\begin{equation}
\tilde{w}_q(u)
=
w_q \exp\!\left(
-\frac{1}{2}
\left(
\frac{\ell_q - \ell_u}{h}
\right)^2
\right),
\end{equation}
where $\ell_u = \log(\max(k_u^{\mathrm{struct}}, 1))$. The node-specific
inflation factor is then the locally weighted mean
\begin{equation}
\hat{c}_u
=
\frac{\sum_q \tilde{w}_q(u)\, r_q}{\sum_q \tilde{w}_q(u)},
\qquad
\text{clamped to } [1,\;\max_q r_q].
\end{equation}
If the pool has no valid spread in structural dimension, the runtime falls
back to the global constant $\hat{c}_u = \hat{c}_{\mathrm{global}}$.

For each focal pair $u$, the adjusted runtime test is
\begin{equation}
T_u^{\mathrm{adj}} = \frac{T_u^{\mathrm{raw}}}{\hat{c}_u},
\qquad
p_u^{\mathrm{sib}}
=
\Pr\!\left(\chi^2_{\nu_u} \ge T_u^{\mathrm{adj}}\right).
\end{equation}
This adjusted test should be read as a practical calibration layer for
post-selection inflation rather than as a
separate finite-sample exact theorem for the adjusted sibling statistic.

The sibling multiplicity correction is ordinary
Benjamini--Hochberg \citep{benjamini1995controlling} at
$\alpha_{\mathrm{sib}} = 0.01$, applied only across focal sibling pairs. Let
$S^{\mathrm{sib}}(u) \in \{0,1\}$ denote the resulting indicator that parent
$u$ supports a significant sibling split.
