% !TEX root = ../../main.tex

\section{Evaluation Questions}

An empirical evaluation of KL-TE should cover four main blocks.

\subsection{Does the Method Stay Calibrated Under the Null?}

The first requirement is a null calibration study on binary data with no true
cluster structure. The goal is to measure how often the child-parent edge test
and the sibling test reject under scenarios where they should not. This should
be reported both at the
individual-test level and at the resulting number-of-clusters level. The latter
is critical because a method can have mild per-test inflation yet still produce
severe over-splitting once the tree walk is applied.

\subsection{Does the Method Recover Planted Structure?}

The second requirement is a controlled power study with planted binary subtrees.
This should vary:

\begin{itemize}[leftmargin=1.5em]
  \item sample size per cluster,
  \item effect size in feature rates,
  \item number of informative features,
  \item tree imbalance, and
  \item background noise level.
\end{itemize}

Power should be measured both for recovery of important tree splits and for the
final clustering outcome. Adjusted Rand Index and exact split recovery are both
useful endpoints.

\subsection{Where Is the Method Robust or Fragile?}

For a paper centered on one implemented method, the key empirical question is
robustness. The most useful stress tests are:

\begin{itemize}[leftmargin=1.5em]
  \item sensitivity to small child leaf counts in the child-parent edge test,
  \item sensitivity to sparse high-dimensional feature matrices,
  \item sensitivity to tree imbalance and shallow-versus-deep signal placement, and
  \item sensitivity of the final partition to the pass-through rule.
\end{itemize}

These analyses should show where the implemented method is well calibrated,
where it becomes under-powered, and where it tends to over-split or
under-split.

\subsection{How Does the Method Behave on Benchmarks?}

The repository already contains benchmark infrastructure and reporting code. A
manuscript-quality benchmark section should align those experiments with the
method claims in the text, focusing on:

\begin{itemize}[leftmargin=1.5em]
  \item synthetic binary benchmarks with known labels,
  \item sensitivity to over-splitting and under-splitting,
  \item comparison to standard clustering baselines, and
  \item qualitative inspection of annotated trees.
\end{itemize}
